\documentclass[sigconf]{acmart}
\usepackage{booktabs}

%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

%\acmDOI{10.475/123_4}

%\acmISBN{123-4567-24-567/08/06}

\acmConference[ELS2018]{European Lisp Symposium}{April 2018}{Marbella,
  Spain}
\acmYear{2018}
\copyrightyear{2018}


\acmArticle{1000}
\acmPrice{1000.00}

%\acmBooktitle{}
%\editor{}


\begin{document}
\title{A perspective of Common Lisp, the UNIX Virtual File System, and
  streams programming}
%\titlenote{}
%\subtitle{}
%\subtitlenote{}


\author{Thomas de Grivel}
%\authornote{}
%\orcid{}
%\affiliation{%
%  \institution{}
%  \streetaddress{}
%  \city{}
%  \state{}
%  \postcode{}
%}
\email{thomasdegrivel@gmail.com}


\begin{abstract}
  This paper gives a few perspectives on Common Lisp and UNIX,
  particularly the Virtual File System with comparisons with HTTP
  REST, and streams programming. These paradigms have been pervasive
  across computer systems for a few decades now and following is a
  review of how the Common Lisp family of languages allows to describe
  and engineer these computing paradigms. They are of prime importance
  in that they give social insight on i/o by privileging hierarchical
  data addressing which remains human readable as much as
  possible. Reviews of the Common Lisp programming and system
  implementation interfaces for hierarchical file naming systems and
  stream programming, comparisons with HTTP REST, the UNIX VFS, and
  Erlang. These are the paradigms that would most likely be needed to
  make Common Lisp relevant in the striving open-source computing
  industry that is moving on to express the world wide web services
  using PHP, SQL, Java, Ruby and Javascript instead.
\end{abstract}


\begin{CCSXML}
  <ccs2012> <concept>
  <concept_id>10011007.10011074.10011075.10011077</concept_id>
  <concept_desc>Software and its engineering~Software design
  engineering</concept_desc>
  <concept_significance>500</concept_significance> </concept>
  <concept>
  <concept_id>10011007.10011006.10011050.10011051</concept_id>
  <concept_desc>Software and its engineering~API
  languages</concept_desc>
  <concept_significance>300</concept_significance> </concept>
  <concept> <concept_id>10010147.10010919.10010177</concept_id>
  <concept_desc>Computing methodologies~Distributed programming
  languages</concept_desc>
  <concept_significance>300</concept_significance> </concept>
  <concept> <concept_id>10002944.10011123.10011673</concept_id>
  <concept_desc>General and reference~Design</concept_desc>
  <concept_significance>100</concept_significance> </concept>
  <concept>
  <concept_id>10003120.10003121.10003124.10010870</concept_id>
  <concept_desc>Human-centered computing~Natural language
  interfaces</concept_desc>
  <concept_significance>100</concept_significance> </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Software design
  engineering} \ccsdesc[300]{Software and its engineering~API
  languages} \ccsdesc[300]{Computing methodologies~Distributed
  programming languages} \ccsdesc[100]{General and reference~Design}
\ccsdesc[100]{Human-centered computing~Natural language interfaces}


\keywords{European Lisp Symposium 2018, Common Lisp, UNIX,
  Hierarchical File Naming Systems, Virtual File Systems, Streams
  Programming}


\maketitle


\section{The UNIX shell and Virtual File System is an acceptable Lisp}

That really is a feeling you get when comming from UNIX and discovering
Common Lisp, that the shell is some kind of low power Lisp. Indeed you
get lists as commands and you need to understand the importance of
quoting things -- interestingly enough there is no QUOTE operator it
is left to be done, often miserably, by the programmer. There is an eval
and you can generate shell scripts using shell commands and send them
to a subshell or pipe them to another shell. So code is data too.
Lists of space separated tokens with special syntax for variable
substitution. The evaluation is mostly command and variable
substitutions on space separated tokens in character streams.
Portability is maximal with all UNIX platforms supported including
Linux and MacOS, on other platforms installation of /bin/sh for
target platform and a few binaries is all you need to do shell
programming.

When you shell script archology is ready you know if it might be useful.
You then just take you small shell scripts one by one and rewrite
them in an actual programming language with a native compiler. So it's
used for rapid prototyping like Lisp languages before a industry
approved solution can come into place.

Where Common Lisp might stand as a valid replacement or complement of
C is that it allows for both rapid prototyping and portable, native,
compiled code. The only limitation then is perception by the industry
regarding Common Lisp as a widely known and widespread and usable
programming language easy to work with, which is not really the case
arguably.

\subsection{An interface to UNIX APIs in Common Lisp : CFFI-POSIX}

Package nicknames in CFFI-POSIX provide UNIX interface names for
standard programming following POSIX specifications. For instance the
CFFI-UNISTD package name gives both implementation details as
CFFI which is a widely available and portable C application binary
interface dynamic linker for Common Lisp, and a POSIX standards
interface name : UNISTD. The package nickname UNISTD only describes
the standard programming interface, because users of the UNISTD
interface may want to change implementation from CFFI- to some other
implementation without changing anything to the source code other than
the system definition files.


\section{How and why is a Hierarchical File Naming System part of a
  computer operating system}

\subsection{Early learning bias}
Hierarchical file naming systems are almost always very early on
produced to students and users of computer systems without their prior
knowledge of it as a necessary interface to persisting data on these
systems. So the concepts of hierarchical naming systems are tightly
linked to persistent on-disk storage and almost identified with the
computer hard drive subsystem very early on. The hard drive subsystem
is one of the slower parts of almost all computer systems so the bias
might have had a negative impact on the development of hierarchical
naming systems as being an essential part of how we describe and
represent structured data as graph cuts for instance when considering
moving, renaming, or processing (mapping) a partial subtree of a
filesystem.

\subsubsection{System implementation versus applicated programming
  interfaces}
As always, APIs (application programming interfaces) are heavily
biased towards concurrency of applications and against concurrency of
systems implementation. This is proeminent in Common Lisp for the
pathname and stream programming interfaces. Most implementations have
found consensus beyond the ANSI Common Lisp language specification
with ``gray streams'' and ``simple streams''.

This bias is somewhat reduced once the concerned subsystem is
expressed in terms of itself. However the hierarchical naming system
seems too dull and simple to express itself in terms of itself. However
one could also say that UNIX succeeded in this by providing its own
VFS program code under the very same filesystem in the kernel, so it
is bundled along with many other powerful abstraction. All UNIX kernel
hacking and filesystem hacking is mainly through a filesystem
demonstrating that large amount of data can be gathered, categorized
and sorted using a virtual file system to the complexity that the
filesystem necessary features can be unambiguously expressed in the
filesystem itself by containing the program code and compiler system
for that code.

Hence the architecture dependent files could only concern the compiler
system. Hierarchical file naming systems and filesystem concepts are
perfectly orthogonal to computing and data processing architecture,
but also allows their representation along with the filesystem itself
using the filesystem.

We argument that usage of hierarchical naming system could definitely
improve the interaction surface of many System and Application
programming interfaces with code editors and software engineers or any
entity operating on symbolic graph cuts (able to do structured I/O).


\subsection{A proposed interface for generic and compatible stream
  programming in Common Lisp : CL-STREAM}

CL-STREAM streams are fully compatible with Common Lisp, but does
shadow most stream related symbols from Common Lisp and redefines
stream primitives constructively as interfaces for both implementor of
stream facilities and application programmers using these streams.


\section{Multi-agent systems and stream programming}

It would seem that almost all simple computing systems can be
effectively made scalable on multi-processing and networked
architectures by replacing list and vector operations by stream
operations. This allows specialization and asynchronous organisation
of processes, auditing, redundancy, auditing, testing and persisting
parts of the intermediate data streams, so it is also important for
finer consistency and reproducibility assesments of a data processing
system.

It follows that success of large scale computer system deployments is
heavily correlated to ability of these systems to rapidly prototype
data processing and then transform them to simple functional
components operating on data streams. The more UNIX and the WWW
progress, the more we see demand for high speed event driven
asynchronous input/output.

Here we review the most commonly deployed systems for multi-agent and
stream programming, namely UNIX and Erlang. These systems readily
expose scalability, parallelization, and distributivity as a direct
consequence of the interfaces they implement and provide to their
users. Next we see how these paradigms can be expressed and engineered
in Common Lisp, eventually overriding Common Lisp stream functions and
classes as being too limited for expressing the full potential of
streams programming.

\subsection{UNIX and C, a structured data streams interface}
UNIX only has binary streams but with C providing casts between data
buffers and structured data, one could say C/UNIX has structured data
streams available directly from the unistd facility and which is
widely used through stdio and sockets. These programming interface are
still widely in use and usually forwarded in some way by programming
languages implemented on these systems.

The following constitute a proposed solution for Common Lisp to stay
on par with most modern operating system facilities to support
structured data and evented asynchronous i/o operations on multiple
streams simultaneously.

\subsubsection{CFFI-POSIX}
Bypassing the need for an update to the language specification and
compiler library, CFFI allows accessing these very fast non-blocking
i/o interfaces from within Common Lisp programs. We present CFFI-POSIX
as a project to maintain a collection of interface to POSIX standards
by the means of CFFI to target all Common Lisp implementations that
are supported by CFFI.

http://github.com/cffi-posix/

\paragraph{CFFI-UNISTD}
provides the standard POSIX UNISTD (ref) interface using the
portable Common Lisp C Foreign Function Interface : CFFI (ref). Included
are direct access to C- functions and Common Lisp wrappers to handle
errors correctly, mostly using CFFI-ERRNO.

http://github.com/cffi-posix/cffi-unistd/

\paragraph{CFFI-FCNTL}
likewise with POSIX FCNTL (ref) interface.

http://github.com/cffi-posix/cffi-fcntl/

\paragraph{CFFI-DIRENT}
likewise with POSIX DIRENT (ref) interface.

http://github.com/cffi-posix/cffi-dirent/

\paragraph{CFFI-SOCKET}
likewise with BSD sockets (ref) interface.

http://github.com/cffi-posix/cffi-socket/

\paragraph{CFFI-ERRNO}
wrappers for POSIX ERRNO (ref) functions.

http://github.com/cffi-posix/cffi-errno/


\subsubsection{Non-blocking streams programming in C/UNIX}
Most high throughput high concurrency data processing systems use
unistd streams or sockets through extensions to address the
limitations of POSIX regarding operation of many open file
descriptors, namely {\em epoll} for Linux and {\em k-queues} for BSD
systems. Good examples of their use are found in recent HTTP servers.

\paragraph{CFFI-EPOLL}
For now only Linux event driven asychronous i/o is supported through the
CFFI-EPOLL package implementing the Linux EPOLL interface.

http://github.com/cffi-posix/cffi-epoll/

\paragraph{THOT}
We know of one webserver that was developped using CFFI-EPOLL and
CFFI-SOCKET to support more than 10k concurrent HTTP connections
on Linux systems : THOT.

http://github.com/RailsOnLisp/thot/


\subsubsection{CL-STREAM}
However we found that the Common Lisp stream programming interfaces
were not sufficient to express non-blocking and buffered i/o that
these CFFI interfaces provide. To address this limitation we
developped CL-STREAM as a mean to describe stream interfaces equally
to application programmers using these interfaces and to systems
implementors writing new way to use or implement these stream
interfaces for non-blocking i/o but also for generic stream
programming in Common Lisp, giving way to all kind of very fast
multi-agent event-driven programming paradigms as a real industry
solution directly in Common Lisp and available in open-source
form. Take it as a standing platform to discuss and describe about all
possible kinds of streams and stream operations in Common Lisp. A
practical minimalistic ontology using CLOS, by all means.

http://github.com/cl-stream/cl-stream/

The CL-STREAM stream interface will accept operations on Common Lisp
streams but do not support non-blocking operations. For those other
implementations have to be devised or found, which is made practical
with the STREAM- generic functions to implement a hierarchy of class
with different stream class traits.

Class hierarchy differenciates between :

\paragraph{streams}
\subparagraph{CLOSE}
when a stream is closed all further i/o operations are prevented from
happening.

\paragraph{input streams}
\subparagraph{READ}
reads one element of type (STREAM-ELEMENT-TYPE stream) from stream or
(STDIN). Return values indicate success, end of file or non-blocking
status.
\subparagraph{STREAM-READ}

\paragraph{buffered input streams}
\subparagraph{MAKE-STREAM-INPUT-BUFFER}
\subparagraph{DISCARD-STREAM-INPUT-BUFFER}
\subparagraph{STREAM-INPUT-BUFFER}
\subparagraph{SETF STREAM-INPUT-BUFFER}
\subparagraph{STREAM-FILL-INPUT-BUFFER}
\subparagraph{STREAM-READ-ELEMENT-FROM-BUFFER}

\paragraph{output streams}
\subparagraph{WRITE}
\subparagraph{STREAM-WRITE}

\paragraph{buffered output streams}
\subparagraph{MAKE-STREAM-OUTPUT-BUFFER}
\subparagraph{CLEAR-OUTPUT}
\subparagraph{STREAM-CLEAR-OUTPUT}
\subparagraph{FINISH-OUTPUT}
\subparagraph{STREAM-FINISH-OUTPUT}
\subparagraph{FLUSH-OUTPUT}
\subparagraph{STREAM-FLUSH-OUTPUT}
\subparagraph{STREAM-OUTPUT-BUFFER}
\subparagraph{SETF STREAM-OUTPUT-BUFFER}
\subparagraph{STREAM-WRITE-ELEMENT-TO-BUFFER}
~\\

This allows fine grained separation of code in packages that each
introduce a minimal number of other systems as dependency requirements.


\subsubsection{UNISTD-STREAM}
UNISTD-STREAM implements the CL-STREAM interface using the UNISTD
interface as provided by CFFI-UNISTD.

http://github.com/cl-stream/unistd-stream/

\subsubsection{BABEL-STREAM}
BABEL-STREAM provides a CL-STREAM interface to support charset encoding
and decoding to make a character stream out of an underlying binary
stream and inversely make a binary stream out of an underlying character
stream. The encoding can be changed on the fly, and the underlying
stream can be read and written too.

http://github.com/cl-stream/babel-stream/

\subsubsection{UNISTD-STDIO}
A replacement for Common Lisp *STANDARD-INPUT*, *STANDARD-OUTPUT*
and *STANDARD-ERROR* using UNISTD-STREAM. Not in working status as of
the writing date of this paper.

http://github.com/cl-stream/babel-stream/

\subsection{Erlang}
All Erlang programs are structured into processes that accept and send
structured data as messages to other processes. So most of the
programming is done by accepting and processing messages sent by other
processes and by sending messages to other processes. Processes can be
pooled, supervised, replicated, distributed across physical hardware
live while the rest of the system keeps running. Structured data
streams, but they call them message queues in Erlang, help programming
multi-agent systems by providing a universal way for os or virtual
machine threads to communicate efficiently.

\subsection{Common Lisp : a minimalistic pathname and streams API}
How does Common Lisp cope beyond specification regarding description
and engineering of hierarchical naming systems and streams programming
?

The main problem is that Common Lisp streams are used to produce the
Common Lisp system, but it is the only implementor of Common Lisp
streams. On modern systems these are actually implemented using the
underlying operating system streams so they inherit the limitations of
the operating system as well as the limitations of Common Lisp
streams. Modern Common Lisp compilers rarely implement Common Lisp
streams in terms of disk I/O and B-tree copy-on-write self balancing
low-level data layers, but rather in terms of kernel syscalls that
implement and persist our human understanding of them through C kernel
code.

So across most Common Lisp implementations the programmer ends up
actually using a maimed and of greater complexity (improved) interface
to the underlying operating system streams. But even with additional
complexity they would have a hard time subclassing and re-implementing
correctly asynchronous operations with just STREAM-READ-CHAR-NO-HANG
as being the only non-blocking operation you can re-define and that is
only if the implementation supports Gray streams. So as a language
specification it really gets in the way of both application programmer
and stream system programmers. Much more than C, which could explain
the difference in adoption rate of these systems to a certain extent,
I would say not zero for sure for anyone except very advanced users.


\
\bibliographystyle{ACM-Reference-Format}
\bibliography{}

\end{document}





- path for newbs ?

- common lisp paper but UNIX heart and soul
  likewise, lisp compiler but UNIX streams

- contexte, raisonnement,

- attractive

- technical points
cl-stream
cffi-posix
